\documentclass[handout]{beamer}
\usepackage[utf8]{inputenc}
\usepackage[T2A]{fontenc}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{array}
\usetheme{Warsaw}
\usecolortheme{wolverine}

\usepackage{listings}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstdefinelanguage{Haskell}%
  {otherkeywords={},%
   morekeywords={as,abstype,if,then,else,case,class,data,default,deriving,%
      hiding,if,in,infix,infixl,infixr,import,instance,let,module,%
      newtype,of,qualified,type,where,do,AbsoluteSeek,AppendMode,%
      Array,BlockBuffering,Bool,BufferMode,Char,Complex,Double,Either,%
      FilePath,Float,Int,Word,Integer,Natural,IO,IOError,Ix,LineBuffering,Maybe,%
      Ordering,NoBuffering,ReadMode,ReadWriteMode,ReadS,RelativeSeek,%
      SeekFromEnd,SeekMode,ShowS,StdGen,String,Void,Bounded,Enum,Eq,%
      Eval,ExitCode,exitFailure,exitSuccess,Floating,Fractional,%
      Functor,Handle,HandlePosn,IOMode,Integral,List,Monad,MonadPlus,%
      MonadZero,Num,Numeric,Ord,Random,RandomGen,Ratio,Rational,Read,%
      Real,RealFloat,RealFrac,Show,System,Prelude,EQ,False,GT,Just,%
      Left,LT,Nothing,Right,WriteMode,True,abs,accum,accumArray,%
      accumulate,acos,acosh,all,and,any,ap,appendFile,applyM,%
      approxRational,array,asTypeOf,asin,asinh,assocs,atan,atan2,atanh,%
      bounds,bracket,bracket_,break,catch,catMaybes,ceiling,chr,cis,%
      compare,concat,concatMap,conjugate,const,cos,cosh,curry,cycle,%
      decodeFloat,delete,deleteBy,deleteFirstsBy,denominator,%
      digitToInt,div,divMod,drop,dropWhile,either,elem,elems,elemIndex,%
      elemIndices,encodeFloat,enumFrom,enumFromThen,enumFromThenTo,%
      enumFromTo,error,even,exitFailure,exitWith,fail,%
      filter,filterM,find,findIndex,findIndices,flip,floatDigits,%
      floatRadix,floatRange,floatToDigits,floor,foldl,foldM,foldl1,%
      foldr,foldr1,fromDouble,fromEnum,fromInt,fromInteger,%
      toInteger,fromJust,fromMaybe,fromRat,fromRational,%
      fromRealFrac,fst,gcd,genericLength,genericTake,genericDrop,%
      genericSplitAt,genericIndex,genericReplicate,getArgs,getChar,%
      getContents,getEnv,getLine,getProgName,getStdGen,getStdRandom,%
      group,groupBy,guard,hClose,hFileSize,hFlush,hGetBuffering,%
      hGetChar,hGetContents,hGetLine,hGetPosn,hIsClosed,hIsEOF,hIsOpen,%
      hIsReadable,hIsSeekable,hIsWritable,hLookAhead,hPutChar,hPutStr,%
      hPutStrLn,hPrint,hReady,hSeek,hSetBuffering,hSetPosn,head,%
      hugsIsEOF,hugsHIsEOF,hugsIsSearchErr,hugsIsNameErr,%
      hugsIsWriteErr,id,ioError,imagPart,indices,init,inits,%
      inRange,insert,insertBy,interact,intersect,intersectBy,%
      intersperse,intToDigit,ioeGetErrorString,ioeGetFileName,%
      ioeGetHandle,isAlreadyExistsError,isAlreadyInUseError,isAlpha,%
      isAlphaNum,isAscii,isControl,isDenormalized,isDoesNotExistError,%
      isDigit,isEOF,isEOFError,isFullError,isHexDigit,isIEEE,%
      isIllegalOperation,isInfinite,isJust,isLower,isNaN,%
      isNegativeZero,isNothing,isOctDigit,isPermissionError,isPrefixOf,%
      isPrint,isSpace,isSuffixOf,isUpper,isUserError,iterate,ixmap,%
      join,last,lcm,length,lex,lexDigits,lexLitChar,liftM,liftM2,%
      liftM3,liftM4,liftM5,lines,listArray,listToMaybe,log,logBase,%
      lookup,magnitude,makePolar,map,mapAccumL,mapAccumR,mapAndUnzipM,%
      mapM,mapM_,mapMaybe,max,maxBound,maximum,maximumBy,maybe,%
      maybeToList,min,minBound,minimum,minimumBy,mkPolar,mkStdGen,%
      mplus,mod,msum,mzero,negate,next,newStdGen,not,notElem,nub,nubBy,%
      null,numerator,odd,openFile,or,ord,otherwise,partition,phase,pi,%
      polar,pred,print,product,properFraction,putChar,putStr,putStrLn,%
      quot,quotRem,random,randomIO,randomR,randomRIO,randomRs,randoms,%
      rangeSize,read,readDec,readFile,readFloat,readHex,readInt,readIO,%
      readList,readLitChar,readLn,readParen,readOct,readSigned,reads,%
      readsPrec,realPart,realToFrac,recip,rem,repeat,replicate,return,%
      reverse,round,scaleFloat,scanl,scanl1,scanr,scanr1,seq,sequence,%
      sequence_,setStdGen,show,showChar,showEFloat,showFFloat,%
      showFloat,showGFloat,showInt,showList,showLitChar,showParen,%
      showSigned,showString,shows,showsPrec,significand,signum,sin,%
      sinh,snd,sort,sortBy,span,split,splitAt,sqrt,stderr,stdin,stdout,%
      strict,subtract,succ,sum,system,tail,tails,take,takeWhile,tan,%
      tanh,toEnum,toInt,toInteger,toLower,toRational,toUpper,transpose,%
      truncate,try,uncurry,undefined,unfoldr,union,unionBy,unless,%
      unlines,until,unwords,unzip,unzip3,unzip4,unzip5,unzip6,unzip7,%
      userError,when,words,writeFile,zero,zip,zip3,zip4,zip5,zip6,zip7,%
      zipWith,zipWithM,zipWithM_,zipWith3,zipWith4,zipWith5,zipWith6,%
      zipWith7},%
   sensitive,%
   morecomment=[l]--,%
   morecomment=[n]{\{-}{-\}},%
   morestring=[b]"%
  }[keywords,comments,strings]%

\lstset{
  language=Haskell,
  showstringspaces=false,
  columns=flexible,
  keepspaces=true,
  basicstyle={\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve}
}

\title{Semilazy data structures in Haskell}
\author[Andrew Lelechenko]{Andrew Lelechenko \\ \texttt{1@dxdy.ru}}
\institute[Barclays]{Barclays, London}
\date{f(by) 2019, Minsk, 26.01.2019}

\begin{document}

\begin{frame}
	\titlepage
\end{frame}

\begin{frame}[fragile]{Be lazy!}

Haskell is a lazy language.
By default the expression (or any of its subexpressions)
is not evaluated until its value
is utterly and unavoidably needed.

\begin{lstlisting}[language=Haskell]
integers :: [Int]
integers = [0..]

f :: Int -> Int
f 42 = 3
f x  = <infinite_loop>

> (map f integers) !! 42
3
\end{lstlisting}

\pause

Laziness is an abstraction to handle potentially infinite processes
as actually infinite objects in a pure functional way.

\end{frame}

\begin{frame}{Introduction to schedules}

A schedule is recursively defined as one of
\begin{itemize}
\item the full calendar,
\pause
\item a literal list of dates,
\pause
\item all specific weekdays (Mondays, Tuesdays, etc.),
\pause
\item union $\cup$ of two schedules,
\pause
\item intersection $\cap$ of two schedules,
\pause
\item all sorts of random stuff,
like each day of a given schedule,
which is the fifth Tuesday of the month
and is directly preceded by a fourth Monday.
\end{itemize}

\bigskip

\pause
It all boils down to the algebra of sets.

\end{frame}

\begin{frame}{A simple schedule}

A trader buys Microsoft stock on New York Stock Exchange
and Dubai Financial Market
and sell it on Moscow Exchange. What is the trading schedule?

\pause
\begin{itemize}
\item
$NYSE$:
  \begin{itemize}
  \item
  Take the full calendar.
  \item
  Remove all Saturdays and Sundays.
  \item
  Remove a list of public holidays in USA.
  \end{itemize}
\pause
\item
$DFM$:
  \begin{itemize}
  \item
  Take the full calendar.
  \item
  Remove all Fridays and Saturdays.
  \item
  Remove a list of public holidays in OAE.
  \end{itemize}
\pause
\item
$MOEX$:
  \begin{itemize}
  \item
  Take the full calendar.
  \item
  Remove all Saturdays and Sundays.
  \item
  Remove a list of public holidays in Russia.
  \item
  Add a list of working holidays in Russia.
  \end{itemize}
\pause
\item
Return $(NYSE \cup DFM) \cap MOEX$.
\end{itemize}


\end{frame}

\begin{frame}{List vs. Set vs. Vector}

How can we represent a set of unique values?

\begin{itemize}
\pause
\item A lazy single-linked list {\tt [a]}: \par
      $O(1)$ insert, $O(n)$ lookup, huge memory overhead.

\pause
\item A fixed-sized array {\tt Vector} {\tt a}: \par
      $O(n)$ insert, $O(\log n)$ lookup, low memory overhead.

\pause
\item A binary search tree {\tt Set} {\tt a}: \par
      $O(\log n)$ insert, $O(\log n)$ lookup, medium memory overhead.
\end{itemize}

\pause
Usually binary search trees are the way to go.

\bigskip

\pause
{\bf Unless the set is infinite.}

\end{frame}

\begin{frame}{How to generate first 100  dates of a schedule?}

How many dates of $x$ and $y$ need to be precomputed
before we are able to return first 100 dates of $x \cap y$?

\bigskip

\pause
We can use try-and-guess: take first 100 dates of both
schedules: $x_{100}$ and $y_{100}$.
If $x_{100} \cap y_{100}$ contains at least 100 dates
we are done. Otherwise take first 200 dates of $x$ and $y$
and try again, etc.
This is ugly and inefficient.

\bigskip

\pause
Another idea is to proclaim a Doomsday
on 29th of February 2900 and compute everything up to this date,
in a vain hope that our system will get decomissioned earlier.
This is vastly inefficient.

\bigskip

\pause
We'd rather work with infinite sets
in a lazy fashion. Unfortunately, we cannot work so with
trees or with arrays.

\pause
\begin{quote}
\ldots when you have eliminated the impossible, whatever remains, however improbable, must be the truth.

\flushright --- One lazy detective
\end{quote}

\end{frame}

\begin{frame}[fragile]{Schedule as a lazy distinct sorted list --- 1}

How to merge (possibly infinite) ordered lists lazily?
We cannot use operations from {\tt Data.List}.

\pause

\begin{lstlisting}[language=Haskell]
merge [1,3,6] [2,4,5,7] = [1,2,3,4,5,6,7]
merge [1,3..] [2,4..]   = [1..]
\end{lstlisting}

\pause
Find inspiration in the merge sort!

\pause
\begin{lstlisting}[language=Haskell]
merge :: Ord a => [a] -> [a] -> [a]
merge [] ys = ys
merge xs [] = xs
merge (x:xs) (y:ys) = case x `compare` y of
    LT -> x : merge xs (y:ys)
    _  -> y : merge (x:xs) ys
\end{lstlisting}

\pause
Similar definitions may be given for set intersection, difference, etc.

\end{frame}

\begin{frame}{Schedule as a lazy distinct sorted list --- 2}

There are several
packages, defining operations on sorted lists:

\begin{table}[h]
\begin{tabular}{ll}
Package & Fatal flaws \\
\hline
{\tt sorted} & NIH, abandoned \\
{\tt sorted-list} & NIH, allows repetitions \\
{\tt data-ordlist} & NIH, provides no type safety \\
\end{tabular}
\end{table}

\pause
{\em Here I intended to insert the XKCD comix about 15 competing standards,
but forgot how to embed pictures in \LaTeX.}

\bigskip

\pause
{\bf Solution:} write a new package {\tt containers-lazy}.
It mimics full {\tt Data.Set} interface,
provides a newtype with safe constructors and
operates over sets without repetitions only.

\bigskip

\pause
Available from {\tt https://github.com/Bodigrim/containers-lazy}

\end{frame}

\begin{frame}[fragile]{Semilazy sets --- 1}

The chosen representation of schedules fits well to listing of first $n$ dates.
Complexity of $\cap$ and $\cup$ is $O(n)$,
same to binary trees.

\medskip

\pause
But lookups suffer
from poor performance: $O(n)$ instead of $O(\log n)$.
Can we make sets great again?

\pause

\begin{lstlisting}[language=Haskell]
data Semilazy a = SL
  { strictInit :: Set a
  ,   lazyTail :: [a]
  }
\end{lstlisting}

E. g., {\tt SL (Set.fromList [1,3,5,9]) [10,20..]}.

\medskip

\pause
{\tt Semilazy} maintains the invariant:
the last element of {\tt strictInit} is less than the first
element of {\tt lazyTail}.
\end{frame}

\begin{frame}[fragile]{Semilazy sets --- 2}

Is this a valid definition of {\tt merge}?

\begin{lstlisting}[language=Haskell]
merge :: Semilazy a -> Semilazy a -> Semilazy a
merge (SL s1 ls1) (SL s2 ls2) =
    SL (s1 `Set.merge` s2) (ls1 `merge` ls2)
\end{lstlisting}

\pause
No, because it does not maintain the invariant:

\begin{lstlisting}[language=Haskell]
merge (SL empty [0..]) (SL (Set.fromList [10]) []) =
   SL (Set.fromList [10] [0..])
\end{lstlisting}

\pause
Valid implementation:

\begin{lstlisting}[language=Haskell]
merge (SL s1 (l1:ls1)) (SL s2 (l2:ls2))
  | l1 < l2, (xs, ys) <- span (< l2) ls1
  = SL (s1 `Set.merge` Set.fromList xs `Set.merge` s2)
       (ys `merge` ls2)
  | otherwise = ...
\end{lstlisting}

\end{frame}

\begin{frame}[fragile]{Semilazy sets --- 3}

To be as lazy as possible
the actual implementation maintains not only a strict init
and a lazy tail, but also a position of delimiter between them.

\begin{lstlisting}[language=Haskell]
data Delimiter a = Bottom | Middle a | Top

data Ascension a = Ascension
    { strictInit :: Set a
    , delimiter  :: Delimiter a
    , lazyTail   :: [a]
    }
\end{lstlisting}

\pause
Available from {\tt https://github.com/Bodigrim/ascension}

\end{frame}

\begin{frame}{Full speed astern}

It is still not entirely satisfying, because lookups take
between $O(\log n)$ and $O(n)$ time.
Can we achieve amortized $O(1)$ time?

\bigskip

\pause
In finite setting when lookups become a bottleneck and inserts are rare,
one can use a bit array. The set is represented by a raw region of memory,
where $i$-th bit equals to 1 when $i$ is an element of the set and equals to 0
otherwise. By the vary nature bit arrays are strict: there is simply no space
to store any pointer to deferred computation.

\bigskip

\pause
Bit arrays provide superfast set intersection / union
by means of bitwise and / or.

\bigskip

\pause
There is a Haskell implementation of bit arrays in {\tt bitvec} package.

\bigskip

\pause
Can we implement an infinite bit array? Since it is infinite
it must somehow involve laziness.

\end{frame}

\begin{frame}{Tricks from a can of worms}

How do dynamic arrays
work in imperative languages? They occupy memory
enough to store $2^k$ elements. While the actual size remains below
$2^k$, appending new elements does not require reallocation.
Only when the size rises beyond $2^k$, new chunk of $2^{k+1}$ size
is allocated and the existing array is copied there.

\bigskip

\pause

Let us have an infinite lazy list of strict bit arrays of growing size:

[ptr to 64 bit block, ptr to 128 bit block, ptr to 256 bit block, ...]

\bigskip

\pause
The lookup function takes an index $n$,
traverses the outer list to extract $m = \log_2 (n/64)$-th element
and returns the relevant bit. For example, to check whether 200 is an
element we traverse until the 3-rd block
and return its $200-(64+128) = 8$-th bit.

\bigskip

\pause
It is better to store bit blocks in a lazy array with instant indexing.
Since chunks grow rapidly, for all practical applications
an outer array of size 64 will suffice. This gives us amortized $O(1)$ indexing.

\end{frame}

\begin{frame}[fragile]{Chimera}

This approach (lazy outer array of pointers to growing inner arrays)
can be generalized from storing bits to storing any data \pause and
is implemented in {\tt chimera} package.

\begin{lstlisting}[language=Haskell]
data Chimera a = Vector (Vector a)
\end{lstlisting}

\pause

{\tt tabulate} takes predicate and returns an infinite bit array:

\begin{lstlisting}[language=Haskell]
tabulate :: (Word -> a) -> Chimera a
\end{lstlisting}

\pause

{\tt index} implements random access in $O(1)$ amortized time:

\begin{lstlisting}[language=Haskell]
index    :: Chimera a -> (Word -> a)
\end{lstlisting}

\end{frame}

\begin{frame}[fragile]{Caching}

Let us use {\tt tabulate} and {\tt index} to get a fully functional caching
in a purely functional and performant manner:

\begin{lstlisting}[language=Haskell]
expensive :: Word -> a
expensive x = <heat_cpu_for_ten_minutes>

cache :: Chimera a
cache = tabulate expensive

cheap :: Word -> a
cheap = index cache
\end{lstlisting}

\end{frame}


\begin{frame}[fragile]{Fibonacci 101}
Let us define Fibonacci numbers in a na\"{\i}ve, exponential way:

\begin{lstlisting}[language=Haskell]
fibo :: Word -> Natural
fibo n = if n < 2 then n else fibo (n-1) + fibo (n-2)
\end{lstlisting}

\pause

We can cache it as is:

\begin{lstlisting}[language=Haskell]
fiboCache :: Chimera Natural
fiboCache = tabulate fibo

fibo' :: Word -> Natural
fibo' = index fiboCache
\end{lstlisting}

But recursive calls still know nothing about cache. Can we make them aware of?

\end{frame}


\begin{frame}[fragile]{Fixed-point combinator}

Any recursive function can be expressed as a non-recursive one and the {\tt fix} combinator
a.~k.~a. Y~combinator.

\begin{lstlisting}[language=Haskell]
fix :: (a -> a) -> a
fix f = let x = f x in x
\end{lstlisting}

\pause

\begin{lstlisting}[language=Haskell]
fiboFix :: (Word -> Natural) -> (Word -> Natural)
fiboFix f n = if n < 2 then n else f (n-1) + f (n-2)

fibo :: Word -> Natural
fibo = fix fiboFix
\end{lstlisting}

\pause
Now use {\tt tabulateFix} to cache all recursive calls as well:

\begin{lstlisting}[language=Haskell]
fiboCache = tabulateFix fiboFix :: Chimera Natural

fibo = index fiboCache :: Word -> Natural
\end{lstlisting}

\end{frame}

\begin{frame}{Summary}

\begin{itemize}
\item Hybrid combination of a strict binary search tree and a lazy list
      allows to work with infinite sets.
\pause
\item If lookups become a bottleneck, one can trade space for speed
      and switch to an infinite bit mask,
      hybrid of a lazy array and a bit array.
\pause
\item The latter approach can be generalized to store any data, applicable
      for transparent caching of functions, including recursive ones.
\end{itemize}

\bigskip
\bigskip

\pause
\centerline{\Huge\bf Thank you!}
\end{frame}

\end{document}
